---
title: "R for Data Science Assessment"
author: "Adeyemo Oluwatosin M."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
  html_notebook:
    toc: yes
    toc_depth: 3
    theme: readable
---
## Introduction
This script contains my solution to analyse the gapminder data set so as to assess my data science skills in R language. The dataset used for this analysis is the gapminder data which contains demographic, health, economic and environmental variables of different countries from 1962 to 2007. The dataset can be found <a href="https://github.com/Bioinformatics-Research-Network/skill-assessments/blob/main/R%20for%20Data%20Science/gapminder_clean.csv"> here </a>

## Import libraries
```{r libraries, warning = FALSE, message = FALSE}
library(tidyverse)
library(ggpubr)
library(plotly)
library(kableExtra)
```

## SECTION A
### Question 1
Read in the data
```{r warning = FALSE, message = FALSE}
gapminder <- read_csv("gapminder_clean.csv")
```

```{r}
gapminder %>%
  sample_n(10) %>%
  kbl() %>%
  kable_classic_2(full_width = F)
```


### Question 2
Filter the data to include only rows where `Year` is `1962` and then make a scatter plot comparing `'CO2 emissions (metric tons per capita)'` and `gdpPercap` for the filtered data.

```{r warning = FALSE, message = FALSE, fig.align="center",fig.width = 10}
plot_1 <- gapminder %>%
  filter(Year == 1962) %>%
  ggplot(mapping = aes(x = `CO2 emissions (metric tons per capita)`, y = gdpPercap)) +
  geom_point() +
  labs(title = "CO2 emmision (metric tons per capita) VS GDP Per capita in 1962", y = "GDP per Capita")

ggplotly(plot_1)
```

### Question 3
On the filtered data, calculate the correlation of `'CO2 emissions (metric tons per capita)'` and `gdpPercap`. What is the correlation and associated p value?

```{r warning = FALSE, message = FALSE}
gapminder %>%
  filter(Year == 1962) %>%
  ggplot(mapping = aes(x = `CO2 emissions (metric tons per capita)`, y = gdpPercap)) +
  geom_point() +
  labs(title = "CO2 emmision (metric tons per capita) VS GDP Per capita in 1962", y = "GDP per Capita") +
  stat_cor()
```

#### Observations

There is a positive correlation between GDP per capital and CO2 emissions in 1962. The R-squared value is 0.93, which means that about 93% variation in GDP per capital can be explained by CO2 emissions. And a p-value that is less than 0.05 shows that the correlation coefficient is statistically significant, meaning that the correlation between the two variables is not by chance. 

### Question 4
On the unfiltered data, answer "In what year is the correlation between `'CO2 emissions (metric tons per capita)'` and `gdpPercap` the strongest?"
```{r warning = FALSE, message = FALSE}
# change the data type of Year from numeric to character
gapminder$Year <- as.character(gapminder$Year)
```


```{r warning = FALSE, message = FALSE}
gapminder %>%
  ggplot(mapping = aes(x = `CO2 emissions (metric tons per capita)`, y = gdpPercap, color = Year)) +
  geom_point() +
  labs(title = "CO2 emissions VS GDP per capital") +
  stat_cor()
```
**The scatter plot shows that 1967 is the year with the strongest correlation between CO2 emissions and GDP per cap**

```{r}
# Filter the data set to 1967
gapminder_1967 <- gapminder %>%
  filter(Year == 1967)
```

### Question 5
Using `plotly`, create an interactive scatter plot comparing `'CO2 emissions (metric tons per capita)'` and `gdpPercap`, where the point size is determined by `pop` (population) and the color is determined by the `continent`. You can easily convert any `ggplot` plot to a `plotly` plot using the `ggplotly()` command.

```{r warning = FALSE, message = FALSE, fig.align="center",fig.width = 10}
plot1 <- ggplot(data = gapminder_1967, mapping = aes(x = `CO2 emissions (metric tons per capita)`, y = gdpPercap, color = continent, size = pop)) +
  geom_point() +
  labs(title = "CO2 emissions VS GDP per capital in 1967")

ggplotly(plot1)
```
#### Observations

1. Asia has the highest CO2 emissions and GDP per capital, and it is also the most populated continent

2. The two variables are positively correlated, an increase in CO2 emissions corresponds to an increase in GDP

## SECTION B

### Question 1
What is the relationship between `continent` and `'Energy use (kg of oil equivalent per capita)'`? (stats test needed)
```{r warning = FALSE, message = FALSE,  fig.align="center",fig.width = 10}
plot2 <- ggplot(data = gapminder, mapping = aes(x = continent, y = `Energy use (kg of oil equivalent per capita)`)) +
  geom_boxplot() +
  scale_y_log10() +
  stat_compare_means(method = "anova") +
  ggtitle("Energy Use Per Continent")

ggplotly(plot2)
```
#### Observations

1. A p-value < 0.05 indicates that there is a significant difference in the energy usage between continents

2. Oceania has the highest median energy use of 3.61 while Africa has the lowest median value with 2.65

### Question 2
Is there a significant difference between Europe and Asia with respect to `'Imports of goods and services (% of GDP)'` in the years after 1990? (stats test needed)

```{r warning = FALSE, message = FALSE}
Gdp_Eur_Asia <- gapminder %>%
  filter(Year > 1990) %>%
  filter(continent == "Europe" | continent == "Asia") %>%
  select(continent, `Imports of goods and services (% of GDP)`)
```

```{r warning = FALSE, message = FALSE,  fig.align="center",fig.width = 10}
plot3 <- ggplot(data = Gdp_Eur_Asia, mapping = aes(x = continent, y = `Imports of goods and services (% of GDP)`)) +
  geom_boxplot() +
  stat_compare_means(method = "t.test") +
  ggtitle(("% of GDP in Europe and Asia"))

ggplotly(plot3)
```
**From the plot above, the p-value > 0.05 which means that there is no significant difference in the % of GDP between Asia and Europe**

### Question 3
What is the country (or countries) that has the highest `'Population density (people per sq. km of land area)'` across all years? (i.e., which country has the highest average ranking in this category across each time point in the data set?)

```{r warning = FALSE, message = FALSE, fig.align="center",fig.width = 10}
plot4 <- gapminder %>%
  select(`Country Name`, Year, `Population density (people per sq. km of land area)`) %>%
  group_by(`Year`, `Country Name`) %>%
  summarise(ave_pop_density = mean(`Population density (people per sq. km of land area)`)) %>%
  arrange(desc(ave_pop_density)) %>%
  slice(1:5) %>%
  ggplot(mapping = aes(x = Year, y = ave_pop_density, color = `Country Name`)) +
  geom_point() +
  labs(title = "Top 5 countries with the highest population density in each year", x = "Year", y = "Population density") +
  theme(axis.text.x = element_text(angle = 50, vjust = 1, hjust = 1, size = 12))

ggplotly(plot4)
```
Five countries with the highest population density across all years are:

1. Monaco

2. Macao SAR, China

3. Singapore

4. Hong kong SAR, China

5. Gibraltar.

### Question 4
What country (or countries) has shown the greatest increase in `'Life expectancy at birth, total (years)'` since 1962? 

```{r warning = FALSE, message = FALSE}
life_expectancy <- gapminder %>%
  select(`Country Name`, `Life expectancy at birth, total (years)`) %>%
  group_by(`Country Name`) %>%
  summarise(ave_life_expectancy = mean(`Life expectancy at birth, total (years)`)) %>%
  arrange(desc(ave_life_expectancy)) %>%
  slice(1:10)

life_expectancy %>%
  kbl(caption = "Top 10 countries with the highest life expectancy") %>%
  kable_classic_2(full_width = F)
```


```{r warning = FALSE, message = FALSE, fig.align="center",fig.width = 10}
plot5 <- ggplot(data = life_expectancy, mapping = aes(x = `Country Name`, y = ave_life_expectancy, color = `Country Name`)) +
  geom_point() +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank()) +
  labs(title = "Countries with the highest life expectancy", y = "Life expectancy")

ggplotly(plot5)
```
**Iceland, Sweden and Japan are the countries with the highest life expectancy**

## Conclusion
From 1962 to 2007, Asia continent is the most populated but Europe (Iceland) has the highest average life expectancy. In all continents, there is a positive correlation between CO2 emission and GDP per capital.
